---
title: "Background"
bibliography: [references.bib, packages.bib]
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#| include: false
knitr::write_bib(
  c("nabor", "mmand", "spatstat.explore", "reticulate"),
  here::here("vignettes/articles/packages.bib")
)
```

## Morphological quantities

Morphological quantities have been studied for a long time already in the
framework of materials research. While this can be traced back to the
determination of effective electrical properties of a dilute suspension of
spherical particles in the first half of the 20th century, only in the 1950s and
1960s, higher-order statistical functions, or so-called n-point probability
functions were introduced, for describing higher concentrations of particles
[see @torquato1982 and references therein]. By the time we started evaluating
morphological quantities of digital representations of material microstructures
(*i.e.* categorical arrays), several "classic" quantities like the two-point
probability function, the lineal path function, and the chord length and pore
size distributions, which are all used until today, were already in place
[@coker1995]. Since then, most research has been related to advective and
diffusive transport in two-phase (*i.e.* pore vs solid) media, focussing on the
relation between their effective transport properties and morphological
quantities [\*e.g.\* @hlushkou2015; @roding2020; @bagherian2021], and the
stochastic reconstruction of random media honouring such morphological
quantities and the corresponding effective properties [\*e.g.\* @yeong1998;
@jiao2009; @pant2015; @lemmens2019a]. More recently, in the late 1990s and early
21st century, with the advent of reactive transport modelling, and need for
increasingly complex phenomenological assessments (*e.g.* in the framework of
radioactive waste disposal), there is an increasing need for morphological
description, as well as stochastic reconstruction, of multi-phase (*i.e.* more
than two) media, where typically different solid phases are distinguished,
additional to the void or pore phase. The **efforts into characterisation of
multi-phase systems have been limited** however [\*e.g.\* @pant2014; @chen2014a;
@lemmens2019a], while recent approaches involving multiple-point statistics,
exemplar-based texture synthesis, and deep learning largely skip the
characterization step, and directly aim at reconstructing [@bostanabad2018].

## Spatial point patterns

While nearest-neighbour statistics have been in use for the description of
particle systems in materials research [\*e.g.\* @torquato1995; @torquato1995a;
@middlemas2020], similar measures of morphology have been in use for the
analysis of spatial point patterns (*e.g.* the reduced second moment or K
function, the pair correlation function, the empty space function), with a wide
range of application domains (from ecology [\*e.g.\* @szmyt2014] to histology
[\*e.g.\* @bull2023]), where they were extended to cross-type measures for
multi-type point patterns (one type vs another, or one type vs all others), and
to 3D point patterns [@baddeley1993]. Instead of applying such cross-type
nearest-neighbour statistics to particle systems or point patterns, they can be
applied to digital representations of microstructures (*i.e.* categorical
arrays) instead. In fact, any categorical array can be seen as a multi-type
point pattern, and application of spatial point pattern analysis [\*e.g.\*
through the tooling provided by @R-spatstat.explore] is straightforward when
using the array indices as spatial coordinates. One important **aspect that is
missing here is the notion of connectivity within a given phase**, which does
come up with important morphological quantities like the lineal path and cluster
function.

## Expectation vs distribution

Most of the classic ways to describe morphology can be based somehow on the
expectation of the amount of neighbouring voxels within a certain distance. Just
**considering the expectation**, or rather a single summary statistic, in fact
means that **we are only extracting a small amount of information from the full
k nearest neighbour distributions for all evaluated points**. We could in fact
also consider measures of spread, or even specific quantiles, to quantify how
consistent the considered descriptor is locally within the assessed array. The
specific morphological quantity labelled "coarseness" has been used for such
purposes [see *e.g.* @coker1995], quantifying variability in the local volume
fraction, in function of a given window size, but there is no reason why we
cannot extend this to all other descriptors where we need to search for more
than a single neighbour. Within the spatial point patterns domain, this is
termed second-order neighbourhood analysis [@getis2010], and gives rise to
localized versions of the commonly used descriptors [see *e.g.*
@R-spatstat.explore], again in function of a given window size.

## The {morphology} R package

The combination of all of the above is exactly the idea behind the {morphology}
package, where we leverage:

1.  The approximate nearest-neighbour search provided by {nabor} [@R-nabor],
    which wraps libnabo, a very performant nearest-neighbour search library
    using a k-d tree data structure [@elseberg2012],
2.  The connected components algorithm provided by {mmand} [@R-mmand], with a
    variable connectivity kernel, and
3.  The shortest path algorithms provided by {igraph},

to implement a flexible API that allows us to

1.  perform different kinds of (k) nearest neighbour queries,
2.  compute all the "classic" morphological quantities, but also very recent
    ones involving shortest paths [@bagherian2021], from such queries by
    applying different kinds of normalisation and using "minus sampling" (aka
    "guard area") edge correction (both used in the spatial point patterns
    [@baddeley1993] and morphological quantities [@coker1995] domains),
3.  compute their cross-category extensions (when applicable, so not for
    connectivity and shortest path-related quantities),
4.  compute 1D, 2D and 3D versions,
5.  enable filtering on 1D, 2D or 3D connectivity,
6.  allow the use of arbitrary summary functions, typically of the numbers of
    neighbours in function of distance.

## Related work

The PoreSpy [@gostick2019] Python package provides similar functionality, but is
constrained a bit closer to the classic morphological quantities. It also has a
broader scope, which includes pore network extraction, exporting to common
formats for process modelling, *etc.*

The MCRpy [@seibert2022c] Python package provides differentiable versions of the
classic morphological quantities, and additionally Gram matrices for
characterizing morphology. The gradient information provided by these can be
leveraged in the reconstruction optimization problem.

While all of this is readily accessible from R through {reticulate}
[@R-reticulate], with {morphology}, we are aiming at more flexible, detailed,
and extensive evaluation of morphological quantities.

## References
