---
title: "Benchmarks"
bibliography: references.bib
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Analytical morphological quantities

Certain morphological quantities can be derived analytically for particular
systems. We consider here those that are evaluated by @pant2016, for fully
penetrable spheres. First we use the fully penetrable spheres generator provided
by the package, and we additionally load and attach the {ggplot2} and {tibble}
packages, to use them later to add the analytical solutions to our plots:

```{r}
library(morphology)
library(ggplot2)
library(tibble)
set.seed(1234)
spheres <- gen_fully_penetrable_spheres()
```

This generates a 200\^3 voxel 3D array by default, with a sphere radius of 15,
exactly like the one used by @pant2016. The result looks as follows:

```{r}
par(mar = rep(0, 4))
spheres[, , 1] |> as.raster() |> plot(interpolate = FALSE)
```

### Two-point correlation function

The two-point correlation function can be obtained by scaling the mean number of
neighbours by the inverse neighbourhood and the proportion. Hence, instead of
providing the mean number of neighbours for a given distance bin, it provides
the probability that voxels with a given separation distance both belong to the
category of interest (here the matrix phase). If we do this for a subset (1/25)
of all x direction 1D subarrays, we can get rather close to the analytical
result derived first by @weissberg1963:

```{r}
v2r <- function(x, r = 15) {
  ifelse(x/r <= 2, 4/3*pi*(1+3/4*x/r-(x/r)^3/16), 8*pi/3)
}
spheres |>
  look_at("voxels", of_category = 0) |> 
  look_in(direction = "x", every = 1/c(5, 1, 5)) |>
  look_for(neighbours = 81, within = 40) |> 
  describe(mean, name = "{morphology}") |> 
  scale_by("inverse neighbourhood", "proportion") |> 
  visualise(x = "distance", y = "probability") +
  geom_line(aes(x, y, colour = "analytical"), data = tibble(
    x = 0:40,
    y = exp(-(392/200^3*15^3)*v2r(x, r = 15))
  ))
```

### Cluster function

The cluster function has as additional requirement that the two voxels are
connected in 3D. Hence, if we repeat the above, but look at `"components"`
instead of `"voxels"`, and ask for describing only the connected cases, we end
up with the cluster function. Unfortunately there is no analytical solution for
this, but since fully penetrable spheres with the currently considered volume
fraction result in a fully connected matrix phase, the cluster function should
be the same as the two-point correlation function, which is checked here:

```{r}
spheres |>
  look_at("components", of_category = 0) |> 
  look_in(direction = "x", every = 1/c(5, 1, 5)) |>
  look_for(neighbours = 81, within = 40, from_border = FALSE) |> 
  describe(mean, connected = TRUE, name = "{morphology}") |> 
  scale_by("inverse neighbourhood", "proportion") |> 
  visualise(x = "distance", y = "probability") +
  geom_line(aes(x, y, colour = "analytical"), data = tibble(
    x = 0:40,
    y = exp(-(392/200^3*15^3)*v2r(x, r = 15))
  ))
```

### Lineal path function

For the lineal path function, we just have to modify the kernel width of the
kernel to determine connectivity, so that we look at 1D connectivity in the
direction of interest. For this, an analytical solution is available from
@lu1992a, and we can reproduce it rather well:

```{r}
spheres |>
  look_at("components", of_category = 0, kernel_width = c(1, 3, 1)) |> 
  look_in(direction = "x", every = 1/c(5, 1, 5)) |>
  look_for(neighbours = 81, within = 40, from_border = FALSE) |> 
  describe(mean, connected = TRUE, name = "{morphology}") |> 
  scale_by("inverse neighbourhood", "proportion") |> 
  visualise(x = "distance", y = "probability") +
  geom_line(aes(x, y, colour = "analytical"), data = tibble(
    x = 0:40,
    y = mean(spheres == 0)^(1 + (3 * x / 4 / 15))
  ))
```

### Chord length distribution

Related to the lineal path function is the chord length distribution, where we
do not look at the probability of two voxels to be connected through a lineal
path, but rather at the distribution of (maximum) lengths of all lineal paths.
This can be done by tabulating (here with base R `table()`) the occurrences of
component IDs in the output from `look_at()` directly, while keeping the kernel
width the same for looking at 1D connectivity. Comparing with the analytical
expression by @torquato1993 gives us:

```{r}
spheres |>
  look_at("components", of_category = 0, kernel_width = c(1, 3, 1)) |>
  _[["component"]] |>
  table() |>
  density(cut = 0, n = 50, bw = 0.5) |> 
  _[c("x", "y")] |>
  transform(name = "{morphology}") |> 
  visualise(x = "chord length", y = "probability") +
  geom_line(aes(x, y, colour = "analytical"), data = tibble(
    x = 0:200,
    y = -3/4/15*log(mean(spheres == 0))*mean(spheres == 0)^(3*x/4/15),
    name = "analytical"  
  ))
```

## Comparison with {spatstat.explore}

### Reduced second moment / K function

If we reduce the size of the dataset, using `every`, then we can more easily (at
least on my machine) compare against the implementation provided by
`spatstat.explore::K3est()` :

```{r}
every <- 1/10
df <- spheres |>
  look_at("voxels", of_category = 0) |> 
  look_in(direction = "xyz", every = every)
rsm_m <- df |>
  look_for(neighbours = 100, within = 20) |> 
  describe(mean, cumulative = TRUE) |> 
  scale_by("inverse proportion", "inverse every")
pp <- spatstat.geom::pp3(df[[1]], df[[2]], df[[3]], rep(c(1, 200), 3))
rsm_s <- spatstat.explore::K3est(pp, rmax = 20, nrval = 21)
as_tibble(rsm_s) |>
  _[, c(1, 3)] |> 
  setNames(c("x", "y")) |>
  transform(name = "{spatstat.explore}") |> 
  rbind(rsm_m |> transform(name = "{morphology}")) |>
  visualise(x = "distance", y = "reduced second moment")
```

### Pair correlation function

It is a bit harder to get a decent match with the results of the implementation
provided by `spatstat.explore::pcf3est()`, as it is using kernel smoothing, but
with adjustments of the default distance in morphology, and playing with the
kernel width adjustment factor, we can get pretty close:

```{r}
every <- 1/5
df <- spheres |>
  look_at("voxels", of_category = 0) |> 
  look_in(direction = "xyz", every = every)
pcf_m <- df |>
  look_for(neighbours = 300, within = 20) |> 
  describe(mean, at = (0:21)-0.5) |> 
  scale_by("inverse neighbourhood", "inverse proportion", "inverse every") 
pp <- spatstat.geom::pp3(df[[1]], df[[2]], df[[3]], rep(c(1, 200), 3))
pcf_s <- spatstat.explore::pcf3est(pp, rmax = 20, nrval = 21, adjust = 0.4, correction = "isotropic")
as_tibble(pcf_s) |>
  _[, c(1, 3)] |> 
  setNames(c("x", "y")) |>
  transform(name = "{spatstat.explore}") |> 
  rbind(pcf_m |> transform(name = "{morphology}")) |>
  visualise(x = "distance", y = "pair correlation function")
```

### Empty space function

As we work here with a regular grid of points instead of a point pattern, the
voxels of our category of interest are matched with themselves, which results in
an offset at distance zero, equal to the proportion of that category (around 0.5
here), for the empty space function. Typically, however, it starts at a
cumulative probability of zero. The implementation provided by
`spatstat.explore::F3est()` does allow for a discrete approximation, however,
and when we set the voxel side length to one for comparing against our
implementation, we can reproduce the {morphology} behaviour:

```{r}
df <- spheres |>
  look_at("voxels", of_category = c(0, 1, 2), in_relation_to = 0) |> 
  look_in(direction = "xyz")
esf_m <- df |>
  look_for(neighbours = 1, within = 20) |> 
  describe(mean, cumulative = TRUE)
df <- df |> subset(value == 2)
pp <- spatstat.geom::pp3(df[[1]], df[[2]], df[[3]], rep(c(1, 200), 3))
esf_s <- spatstat.explore::F3est(pp, rmax = 20, nrval = 21, vside = 1)
as_tibble(esf_s) |>
  _[, c(1, 3)] |> 
  setNames(c("x", "y")) |>
  transform(name = "{spatstat.explore}") |> 
  rbind(esf_m |> transform(name = "{morphology}")) |>
  visualise(x = "distance", y = "cumulative probability")
```

## Comparison with {gstat}

### Variogram

For the (indicator) variogram, the quickest solution is to look into
cross-category pairs here. Every distance bin becomes a Bernouilli variable,
where the squared difference is 1 in case of different categories, and 0 when
they are the same. The probability of a difference of 1 is equal to double the
probability of 0-1 pairs (as this is the same as that for 1-0 pairs), and the
probability of a difference of 0 is then the complement of that. With a final
division by two to get to the semivariance, we get:

```{r}
df <- spheres[1:40, 1:40, 1:40] |> reshape2::melt() |> as_tibble()
g <- gstat::gstat(
  id = "phase",
  formula = value ~ 1,
  locations = ~ Var1 + Var2 + Var3,
  data = df
)
vgm_g <- gstat::variogram(g, cutoff = 20, width = 1) |>
  transform(x = dist, y = gamma, name = "{gstat}") |>
  _[c("x", "y", "name")]
vgm_m <- spheres |>
  look_at("voxels", of_category = 0, in_relation_to = 1) |> 
  look_in(direction = "x") |> 
  look_for(neighbours = 41, within = 20, from_border = TRUE) |> 
  describe(mean) |>
  scale_by("inverse neighbourhood", "proportion") |>
  transform(name = "{morphology}")
vgm_m$y <- (2 * vgm_m$y) * (1 - 2 * vgm_m$y) / 2
visualise(rbind(vgm_m, vgm_g), x = "distance", y = "semivariance")
```

## References
